Mental simulation as Bayesian Quadrature

### 2013-05-31 man/nips-2013.tex
Add TODO
Some paper cleanup

@@

Consider the objects in Figure \ref{fig:mental-rotation}. In each
panel, are the two depicted objects identical (except for a rotation),
or distinct? When presented with this mental rotation task, people
default to a strategy in which they visualize one object rotating
until it is congruent with the other \cite{Shepard1971}. There is
strong evidence for such \textit{mental imagery} or \textit{mental
  simulation}: we can imagine three-dimensional objects in our minds
and manipulate them, to a certain extent, as if they were real
\cite{Kosslyn:2009tj}.  However, the use of mental simulation is
predicated on determining appropriate parameters to give the
simulation, and people's cognitive constraints may furthermore place
limitations on the duration or precision of simulation. One hypothesis
for how these issues are handled argues that people use a rational
solution, meaning that it is optimal under existing constraints
\cite{Lieder:2012wg,Vul:2009wy,Griffiths2012a}, such as processing
speed, memory, or perceptual acuity.

@@

There is a long and rich history of work on mental rotation beginning
with \cite{Shepard1971}, who presented participants with pairs of
images such as those shown in Figure \ref{fig:mental-rotation}. These
pairs of three-dimensional objects could either be: the same object
rotated in plane (Figure \ref{fig:mental-rotation}A), the same object
rotated in depth (Figure \ref{fig:mental-rotation}B), or two different
objects (Figure \ref{fig:mental-rotation}C; in this case, the
``different'' objects had reflective symmetry, but no rotational
symmetry). Participants had to determine whether the two images were
of the same object or not, and \cite{Shepard1971} famously found that
response times for plane and depth rotations had a strong linear
relationship with the minimum angle of rotation between the two
objects. The conclusion was that participants were visually
``rotating'' the objects in their minds.

@@

Because the vertices are connected in a way which forms a closed loop,
we need only consider $n$ mappings, $M$, of the $n$ vertices (we
assume the uncertainty is only in which is the ``first''
vertex). Thus, the possible orderings are of the form $M=\lbrace{}0,
1, \ldots{}, n\rbrace{}$, $M=\lbrace{}n, 0, \ldots{}, n-1\rbrace{}$,
and so on. This gives us an explicit form for the similarity function:
\begin{equation}
  S(X_b, X_R)=\frac{1}{n}\sum_{M\in\mathbb{M}}\prod_{i=1}^n\mathcal{N}(X_b[i]\ \vert \ (MX_R)[i], \Sigma)
  \label{eq:similarity}
\end{equation}
where $i$ denotes the $i^{th}$ vertex. An example of $S$ for the
stimuli shown in Figure \ref{fig:stimuli} is illustrated in Figure
\ref{fig:likelihood}.

@@

\subsection{Gold standard}

@@

\subsection{\Naive{} model}

@@

\subsection{Parametric (Von Mises) model}

@@

\subsection{Nonparametric (Bayesian Quadrature) model}

@@

Bayesian Quadrature has its difficulties, however. While in our case
$S$ is a non-negative likelihood function, GP regression enforces no
such constraint. In an effort to avoid this problem,
\cite{Osborne:2012tm} give a method which involves instead placing a
prior over the log likelihood\footnote{In practice, as in
  \cite{Osborne:2012tm}, we use the transform of
  $\log(S+1)$. Additionally, while \cite{Osborne:2012tm} use a
  combination of MLII and marginalization to fit the kernel
  parameters, we set the output scale $h_S=0.125$ and $h_{\log
    S}=\log(h_S + 1)$, and use MLII to fit all other parameters.},
thus ensuring that $S=e^{\log S}$ will be positive\footnote{We are not
  guaranteed positivity, however, the approximation to the integral
  over $\log S$ (Equation \ref{eq:bq-Z-mean}) requires computing
  $\mu_S$, which may have non-positive segments.}:
\begin{equation*}
  E[Z\ \vert \ \log S]=\int_{\log S}\left(\int_R \exp(\log{S(X_b,X_R)})p(R)\ \mathrm{d}R\right)\mathcal{N}\left(\log{S}\ \vert \ \mu_{\log S}, \Sigma_{\log S}\right)\ \mathrm{d}\log S
\end{equation*}
where $\mu_{\log S}$ and $\Sigma_{\log S}$ are the mean and
covariance, respectively, of the GP regression over $\log S$ given
$\mathbf{R}$. We approximate this according to the method given in
\cite{Osborne:2012tm}:
\begin{equation}
  \mu_Z=E[Z\ \vert \ S, \log S, \Delta_c] \approx \int_R \mu_{S}(1 + \mu_{\Delta_c}) p(R)\ \mathrm{d}R 
  \label{eq:bq-Z-mean}
\end{equation}
where $\mu_S$ is the mean of a GP regression over $S$ given
$\mathbf{R}$; and $\mu_{\Delta_c}$ is a regression over
$\Delta_c=\mu_{\log S} - \log \mu_S$ given $\mathbf{R}_c$, which
consists of $\mathbf{R}$ and a set of intermediate \emph{candidate
  points}\footnote{These candidate points do not require evaluating
  the true $S$, only the GP estimates of $S$ and $\log S$.} $c$ as
described in \cite{Osborne:2012tm}. The variance of the estimate of
$Z$ is given by:
\begin{equation}
  \sigma_Z=\mathrm{Var}(Z\ \vert \ S, \log S, \Delta_c) = \int_R\int_{R^\prime} \mu_S(R)\mu_S(R^\prime) \Sigma_{\log S}(R, R^\prime)p(R)p(R^\prime)\ \mathrm{d}R\ \mathrm{d}R^\prime
  \label{eq:bq-Z-var}
\end{equation}

To choose $\mathbf{R}$, we pick an initial direction of rotation which
results in the higher value of $S$, and then continue rotating in that
direction until the variance in Equation \ref{eq:bq-Z-var} is low
enough that we are confident about the likelihood ratio $\ell$:
\begin{equation*}
p(\ell)\approx\frac{1}{p(X_b)}\ \mathcal{N}(Z\ \vert\ \mu_Z, \sigma_z)
\end{equation*}
Specifically, we choose $h_0$ when $p(\ell < 1)\geq 0.95$, and we
choose $h_1$ when $p(\ell > 1)\geq 0.95$. Until one of these
conditions are met (or the shape has been fully rotated), the model
will continue to compute rotations and update its estimate of $Z$.

@@

We evaluated each model's performance on 50 randomly generated shapes
(such as $X_a$ in Figure \ref{fig:stimuli}). For each shape, we picked
a random angle $R$ and generated two stimuli: a $h_0$ stimulus, by
reflecting $X_a$ across the $y$-axis and rotating it by $R$; and a
$h_1$ stimulus, by simply rotating $X_a$ by $R$.

We considered three metrics of performance in particular:
\begin{itemize}
\item \textit{Rotations}: for stimuli where $h_1$ is correct, how
  correlated are the true angles of rotation with the rotations
  performed by the model? This is defined to be the Pearson's
  correlation coefficient $\rho$ between $R$ and $\vert
  \mathbf{R}\vert$. Figure \ref{fig:rotations} shows individual points
  corresponding to true rotation vs. rotation by the model, for each
  model.
\item \textit{Responses}: how accurate is the model at choosing the
  correct hypothesis? This is defined to be the mean error ($\ME{}$),
  or fraction of times the model picks the incorrect hypothesis.
\item \textit{Estimates of $Z$}: how accurate is the model's estimate
  of $Z$? This is defined to be the mean squared error ($\MSE{}$)
  between the model's estimate of $Z$ and the ``gold standard'' value,
  where the error has scale such that $\MSE{}=0$ indicates no error,
  and $\MSE{}=1$ indicates maximum error. Figure \ref{fig:accuracy}
  shows plots of the true (gold standard) value of $Z$ vs. the model's
  estimate, for each model.
\end{itemize}

>>>>

\TODO{Want to talk about previous work on strategies people might use,
  e.g. eye-tracking results from Just and Carpenter (1975?)}

@@

Consider the images on the left in Figure
\ref{fig:mental-rotation}. In each panel, are the two depicted objects
identical (except for a rotation), or distinct? When presented with
this mental rotation task, people default to a strategy in which they
visualize one object rotating until it is congruent with the other
\cite{Shepard1971}. There is strong evidence for such \textit{mental
  imagery} or \textit{mental simulation}: we can imagine
three-dimensional objects in our minds and manipulate them, to a
certain extent, as if they were real \cite{Kosslyn:2009tj}.  However,
the use of mental simulation is predicated on determining appropriate
parameters to give the simulation, and people's cognitive constraints
may furthermore place limitations on the duration or precision of
simulation. One hypothesis for how these issues are handled argues
that people use a rational solution, meaning that it is optimal under
existing constraints \cite{Lieder:2012wg,Vul:2009wy,Griffiths2012a},
such as processing speed, memory, or perceptual acuity.

@@

There is a long and rich history of work on mental rotation beginning
with Shepard and Metzler \cite{Shepard1971}, who presented
participants with pairs of images such as those shown in Figure
\ref{fig:mental-rotation} (left panel). These pairs of
three-dimensional objects could either be: the same object rotated in
plane (Figure \ref{fig:mental-rotation}A), the same object rotated in
depth (Figure \ref{fig:mental-rotation}B), or two different objects
(Figure \ref{fig:mental-rotation}C; in this case, the ``different''
objects had reflective symmetry, but no rotational
symmetry). Participants had to determine whether the two images were
of the same object or not, and \cite{Shepard1971} famously found that
response times for plane and depth rotations had a strong linear
relationship with the minimum angle of rotation between the two
objects (Figure \ref{fig:mental-rotation}, right panel). The
conclusion was that participants were visually ``rotating'' the
objects in their minds.

@@

Because the vertices are connected in a way which forms a closed loop,
we need only consider $2n$ mappings, $M$, of the $n$ vertices (we
assume uncertainty for which is the ``first'' vertex, and then which
of its two neighbors is the ``second''). Thus, the possible orderings
are of the form $M=\lbrace{}0, 1, \ldots{}, n\rbrace{}$,
$M=\lbrace{}n, 0, \ldots{}, n-1\rbrace{}$, and so on. This gives us an
explicit form for the similarity function:
\begin{equation}
  S(X_b, X_R)=\frac{1}{2n}\sum_{M\in\mathbb{M}}\prod_{i=1}^n\mathcal{N}(X_b[i]\ \vert \ (MX_R)[i], \Sigma)
  \label{eq:similarity}
\end{equation}
where $i$ denotes the $i^{th}$ vertex. An example of $S$ for the
stimuli shown in Figure \ref{fig:stimuli} is illustrated in Figure
\ref{fig:likelihood}.

@@

@@

@@

@@

@@

Bayesian Quadrature has its difficulties, however. While in our case
$S$ is a non-negative likelihood function, GP regression enforces no
such constraint. In an effort to avoid this problem,
\cite{Osborne:2012tm} give a method which involves instead placing a
prior over the log likelihood\footnote{In practice, as in
  \cite{Osborne:2012tm}, we use the transform of
  $\log(S+1)$. Additionally, while \cite{Osborne:2012tm} use a
  combination of MLII and marginalization to fit the kernel
  parameters, we set the output scale $h_S=\sqrt{0.1}$ and $h_{\log
    S}=\log(h_S + 1)$, and use MLII to fit all other parameters.},
thus ensuring that $S=e^{\log S}$ will be positive\footnote{We are not
  guaranteed positivity, however, the approximation to the integral
  over $\log S$ (Equation \ref{eq:bq-Z-mean}) requires computing
  $\mu_S$, which may have non-positive segments.}:
\begin{equation*}
  E[Z\ \vert \ \log S]=\int_{\log S}\left(\int_R \exp(\log{S(X_b,X_R)})p(R)\ \mathrm{d}R\right)\mathcal{N}\left(\log{S}\ \vert \ \mu_{\log S}, \Sigma_{\log S}\right)\ \mathrm{d}\log S
\end{equation*}
where $\mu_{\log S}$ and $\Sigma_{\log S}$ are the mean and
covariance, respectively, of the GP regression over $\log S$ given
$\mathbf{R}$. We approximate this according to the method given in
\cite{Osborne:2012tm}:
\begin{equation}
  \mu_Z=E[Z\ \vert \ S, \log S, \Delta_c] \approx \int_R \mu_{S}(1 + \mu_{\Delta_c}) p(R)\ \mathrm{d}R 
  \label{eq:bq-Z-mean}
\end{equation}
where $\mu_S$ is the mean of a GP regression over $S$ given
$\mathbf{R}$; and $\mu_{\Delta_c}$ is a regression over
$\Delta_c=\mu_{\log S} - \log \mu_S$ given $\mathbf{R}_c$, which
consists of $\mathbf{R}$ and a set of intermediate \emph{candidate
  points}\footnote{These candidate points do not require evaluating
  the true $S$, only the GP estimates of $S$ and $\log S$.} $c$ as
described in \cite{Osborne:2012tm}. The variance of the estimate of
$Z$ is given by:
\begin{equation}
  \sigma_Z=\mathrm{Var}(Z\ \vert \ S, \log S, \Delta_c) = \int_R\int_{R^\prime} \mu_S(R)\mu_S(R^\prime) \Sigma_{\log S}(R, R^\prime)p(R)p(R^\prime)\ \mathrm{d}R\ \mathrm{d}R^\prime
  \label{eq:bq-Z-var}
\end{equation}

\TODO{To choose $\mathbf{R}$}, we pick an initial direction of
rotation which results in the higher value of $S$, and then continue
rotating in that direction until the variance in Equation
\ref{eq:bq-Z-var} is low enough that we are confident about the
likelihood ratio $\ell$:
\begin{equation*}
p(\ell)\approx\frac{1}{p(X_b)}\ \mathcal{N}(Z\ \vert\ \mu_Z, \sigma_z)
\end{equation*}
Specifically, we choose $h_0$ when $p(\ell < 1)\geq 0.95$, and we
choose $h_1$ when $p(\ell > 1)\geq 0.95$. Until one of these
conditions are met (or the shape has been fully rotated), the model
will continue to compute rotations and update its estimate of $Z$.

@@

We evaluated each model's performance on 20 randomly generated shapes
which had between 3 and 6 vertices, inclusive, (e.g. $X_a$ in Figure
\ref{fig:stimuli}). For each shape, we computed 18 ``different'' and
18 ``same'' stimuli pairs, with $R$ spaced at $20^\prime$ increments
between 0 and 360. ``Different'' pairs were created by reflecting
$X_a$ across the $y$-axis and rotating it by $R$; for ``same'' pairs,
$X_a$ was rotated by $R$.

We considered three metrics of performance in particular:
\begin{itemize}
\item \textit{Rotations}: for ``same'' pairs, how correlated are the
  true angles of rotation with the rotations performed by the model?
  This is defined to be the Pearson's correlation coefficient $\rho$
  between $R$ and $\vert \mathbf{R}\vert$. Figure \ref{fig:rotations}
  shows individual points corresponding to true rotation vs. rotation
  by the model, for each model.
\item \textit{Responses}: how accurate is the model at choosing the
  correct hypothesis? This is defined to be the mean error ($\ME{}$),
  or fraction of times the model picks the incorrect hypothesis.
\item \textit{Estimates of $Z$}: how accurate is the model's estimate
  of $Z$? This is defined to be the mean squared error ($\MSE{}$)
  between the model's estimate of $Z$ and the ``gold standard'' value,
  where the error has scale such that $\MSE{}=0$ indicates no error,
  and $\MSE{}=1$ indicates maximum error. Figure \ref{fig:accuracy}
  shows plots of the true (gold standard) value of $Z$ vs. the model's
  estimate, for each model.
\end{itemize}
